Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-29 text/back_end_KS.qmd
DFW smoothing results

From 1980 to 2020, the population of the overall Dallas-Fort Worth metropolitan area increased substantially from approximately 2.7 Million to 6.8 Million. @fig-dallas-maps showcases all six Geomarkets experienced population growth over the three census periods. However, the most pronounced increases occured in the West of Dallas/Fort Worth Metroplex and Collin & Rockwall Counties Geomarkets, which grew from approximately 340,000 to 1.4 Million and 170,000 to 1.2 Million, respectively. These population trends, as shown in @fig-dallas-maps, suggest population growth has been partly driven by urban sprawl from the metropolitan centers - City of Forth Worth and City of Dallas - into the surrounding suburban and rural areas of the metropolitan areas. This outward expansion may also account for the comparatively larger geographic size of the West of Dallas/Fort Worth Metroplex Geomarket.

@fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Ft. Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County (58% White, 16% Black, 19% Hispanic; 6% Asian Pacific Islander), Irving, Arlington, and Grand Prairie (54% White, 12% Black, 26% Hispanic; 16% Asian Pacific Islander), and City of Fort Worth (57% White, 16% Black, 22% Hispanic; 3% Asian Pacific Islander). In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively. Notably, by 2000, the City of Dallas was no longer a predominantly White Geomarket, with a racial/ethnic composition of approximately 24% Black, 30% Hispanic, and 3% Asian Pacific Islander residents.

@@

@fig-dallas-rq1-ses also illustrates changes in the percentage of residents living in poverty within Geomarkets over time, revealing shifting dynamics across Geomarkets similar to those observed in median household income. In 1980, poverty rates were comparably low in both Collin & Rockwall Counties and Dallas County, moderate in the Irving, Arlington, and Grand Prarie and West of Dallas/Fort Worth Metroplex Geomarkets, and highest for City of Dallas and City of Fort Worth Geomarkets. Overtime, poverty rates declined in Collin & Rockwall Counties and West of Dallas/Ft. Worth Metroplex, while increasing for all other Geomarkets. These divergent trends resulted in a widening gap in poverty levels across the Dallas-Fort Worth metropolitan area. A similar widening disparity is evident evident in educational attainment. Although all Geomarkets experienced increases in the percentage of residents with a bachelor's degree or higher between 1980 and 2020, the Collin & Rockwall Counties Geomarket experienced a substantially greater increase - from 23% in 1980 to 50% in 2020- maintaining the gap with other Geomarkets in terms of educational attainment.

>>>>

From 1980 to 2020, the population of the overall Dallas-Fort Worth metropolitan area increased substantially from approximately 2.7 Million to 6.8 Million. @fig-dallas-maps showcases all six Geomarkets experienced population growth over the three census periods. However, the most pronounced increases occured in the West of Dallas/Fort Worth Metroplex and Collin & Rockwall Counties Geomarkets, which grew from approximately 340,000 to 1.4 Million and 170,000 to 1.2 Million, respectively. These population trends, as shown in @fig-dallas-maps, suggest population growth has been partly driven by urban sprawl from the metropolitan centers - City of Forth Worth and City of Dallas - into the surrounding suburban and rural areas of the metropolitan areas. This outward expansion may also account for the comparatively larger geographic size of the West of Dallas/Fort Worth Metroplex Geomarket from 1980 to 2000.

@fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Fort Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County (58% White, 16% Black, 19% Hispanic; 6% Asian Pacific Islander), Irving, Arlington, and Grand Prairie (54% White, 12% Black, 26% Hispanic; 16% Asian Pacific Islander), and City of Fort Worth (57% White, 16% Black, 22% Hispanic; 3% Asian Pacific Islander). In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively. Notably, by 2000, the City of Dallas was no longer a predominantly White Geomarket, with a racial/ethnic composition of approximately 24% Black, 30% Hispanic, and 3% Asian Pacific Islander residents.

@@

@fig-dallas-rq1-ses also illustrates changes in the percentage of residents living in poverty within Geomarkets over time, revealing shifting dynamics across Geomarkets similar to those observed in median household income. In 1980, poverty rates were comparably low in both Collin & Rockwall Counties and Dallas County, moderate in the Irving, Arlington, and Grand Prarie and West of Dallas/Fort Worth Metroplex Geomarkets, and highest for City of Dallas and City of Fort Worth Geomarkets. Overtime, poverty rates declined in Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex, while increasing for all other Geomarkets. These divergent trends resulted in a widening gap in poverty levels across the Dallas-Fort Worth metropolitan area. A similar widening disparity is evident in educational attainment. Although all Geomarkets experienced increases in the percentage of residents with a bachelor's degree or higher between 1980 and 2020, the Collin & Rockwall Counties Geomarket experienced a substantially greater increase - from 23% in 1980 to 50% in 2020- maintaining the gap with other Geomarkets in terms of educational attainment.

The Dallas-Fort Worth metroplex experienced substantial demographic growth since Geomarket boundaries were dilineated by @RN4982 in the early 1980s through 2020. The metropolitan area's unique urban structure of a dual central cities surrounded by concentric suburbs has shaped patterns of population growth across racial/ethnic, economic, and spatial shifts in the Geomarket lanscape. While all six Geomarkets experienced population increases, the most significant growth occurred in the outer suburban areas, reflecting ongoing urban sprawl. However, all Geomakets have experienced substantial growth in Hispanic, Black, and Asian populations leading to only two of the six Dallas-Fort Worth Geomarkets remaining predominantly White (Collin and Rockwall Counties and West of Dallas/Fort Worth Metroplex. On the other hand, the metropolitan area has experienced widening disparities in income, poverty, and education attainment across Geomarkets, especially between the historically affluent Collin and Rockwall Counties and the less economically advantaged inner-core Geomarkets (City of Fort Worth and City of Dallas). Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.
